---
title: "Part 2 Basic Inferential Data Analysis"
author: "John B Cheadle"
date: "August 10, 2017"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(datasets)
```


## Overview
This report describes analysis of the ToothGrowth data in the R datasets package.  Here we load and summarize the data using basic exploratory analysis, then compare tooth growth by supp and dose using confidence intervals and hypothesis tests.  

*Did the student perform some relevant confidence intervals and/or tests?

*Were the results of the tests and/or intervals interpreted in the context of the problem correctly?

*Did the student describe the assumptions needed for their conclusions?

## ToothGrowth Data
We load the ToothGrowth data frame from the R datasets package and explore some of its characteristics.
```{r tg_data, echo=TRUE}
tg <- ToothGrowth
dim(tg)
head(tg)
```
Briefly it appears that there are two supplements (supp), 'VC" and "OJ', with doses of 0.5, 1.0, and 2.0 for each.  'len' looks to be a measurement of tooth growth length for each of these groups.

```{r tg_plots, echo=TRUE}
## Split data into supplements
VC_tg <- tg[tg$supp == "VC",]
OJ_tg <- tg[tg$supp == "OJ",]

## Plot boxplots of TG data for each supp
par(mfrow=c(1,2))
boxplot(formula=len ~ dose, data=VC_tg,names=unique(VC_tg$dose), main="supp: VC", xlab="dose", ylab="len")
boxplot(formula=len ~ dose, data=OJ_tg,names=unique(OJ_tg$dose), main="supp: OJ", xlab="dose", ylab="len")
```

## Summary of ToothGrowth Data
We see a linear relationship between dose and len for both supplement groups (VC and OJ).   

## Comparison of tooth growth by supp and dose

### Confidence Intervals

### Hypothesis Tests

## Conclusions
What assumptions did I need to use here?